### 2024-06-28 paper/tex/04-algorithm.tex
edit

In this section, we present the algorithmization of the declarative system described above.
The algorithmic system follows the structure of the declarative specification closely.
First, it is also given by a set of inference rules, which, however,
are mode-correct (\cite{dunfield2020:bidirectional}), \ie
the output of each rule is always uniquely defined by its input.
And second, most of the declarative rules 
(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel})
have a unique algorithmic counterpart, 
which simplifies reasoning about the algorithm and its correctness proofs.

@@

First, let us discuss the syntax of the algorithmic system.

@@

\paragraph*{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

@@

\subsection{Type Constraints}
Throughout the algorithm's operation, it gathers information about the
algorithmic type variables, represented as \emph{constraints}. These constraints
in our system can be either \emph{subtyping constraints} or \emph{unification
constraints}. As preserved by the algorithm, each subtyping constraint has a
positive shape $[[α̂⁺ :≥ iP]]$, meaning it confines a positive algorithmic
variable to be the supertype of a positive declarative type. Unification
constraints can take a positive $[[α̂⁺ :≈ iP]]$ or negative $[[α̂⁻ :≈ iN]]$
form, but algorithmic type variables cannot occur on their right-hand side. The
constraints \emph{set} is denoted as $[[SC]]$, and we presume that each
algorithmic variable can be restricted by at most one constraint.

We define $[[UC]]$ separately as a set solely containing \emph{unification}
constraints for simpler algorithm representation. The unification algorithm,
which we use as a subroutine of the subtyping algorithm, can only produce
unification constraints. A resolution of unification constraints is simpler than
that of a general constraint set. This way, this separation allows us to better
decompose the algorithm's structure, thus simplifying the inductive proofs.

@@

\paragraph*{Auxiliary Functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

@@

\paragraph*{Constraint well-formedness and satisfaction}
Suppose that $[[Θ]]$ is an instantiation context. 
We say that constraint set $[[SC]]$ is well-formed in $[[Θ]]$ 
(denoted as $[[Θ ⊢ SC]]$) 
if for every every constrain entry $[[scE ∊ SC]]$
associating a variable $[[α̂±]]$ with a type $[[iP]]$,
this type is well-formed in the corresponding declarative context $[[Θ(α̂±)]]$.

>>>>

In this section, we present the algorithmization of the declarative system described above.
The algorithmic system follows the structure of the declarative specification closely.
First, it is also given by a set of inference rules, which, however,
are mode-correct (\cite{dunfield2020:bidirectional}), \ie
the output of each rule is always uniquely defined by its input.
And second, the rules of the algorithmic system mirrors the declarative rules 
(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel}), 
which simplifies the correctness proof.

@@

First, let us discuss the syntax of the algorithmic system (\cref{fig:algo-syntax}).

@@

\paragraph*{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context $[[Γ]]$, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

@@

\subsection{Type Constraints}
Throughout the algorithm's operation, it gathers information about the
algorithmic type variables, represented as \emph{constraints}. These constraints
in our system can be either \emph{subtyping constraints} or \emph{unification
constraints}. As preserved by the algorithm, each subtyping constraint has a
positive shape $[[α̂⁺ :≥ iP]]$, meaning it confines a positive algorithmic
variable to be the supertype of a positive declarative type. Unification
constraints can take a positive $[[α̂⁺ :≈ iP]]$ or negative $[[α̂⁻ :≈ iN]]$
form. Algorithmic variables cannot occur in the right-hand side of the
constraints.  The constraint \emph{set} is denoted as $[[SC]]$, and we presume that each
algorithmic variable can be restricted by at most one constraint.

We define $[[UC]]$ separately as a set solely containing \emph{unification}
constraints for simpler algorithm representation. The unification algorithm,
which we use as a subroutine of the subtyping algorithm, can only produce
unification constraints. The resolution of unification constraints is simpler than
that of a general constraint set. This way, this separation allows us to better
decompose the algorithm's structure, thus simplifying the inductive proofs.

@@

\paragraph*{Auxiliary Functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context $[[Θ]]$---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

@@

\paragraph*{Constraint well-formedness and satisfaction}
Suppose that $[[Θ]]$ is an instantiation context. 
We say that constraint set $[[SC]]$ is well-formed in $[[Θ]]$ 
(denoted as $[[Θ ⊢ SC]]$) 
if for every constrain entry $[[scE ∊ SC]]$
associating a variable $[[α̂±]]$ with a type $[[iP]]$,
this type is well-formed in the corresponding declarative context $[[Θ(α̂±)]]$.
